---
title: "Principal Component Analysis: Application and Implementation"
author: "Aklanta Borah, Amartya Amritanshu, Rajveer Vora, Somesh Biswas"
date: "30 April, 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

\newpage

# PCA on the Iris Dataset

## The Dataset

The Iris dataset consists of **150 observations** from 3 species of flowers (*setosa*, *versicolor*, *virginica*), each measured across 4 features:

-   $x_1$: Sepal Length (cm)
-   $x_2$: Sepal Width (cm)
-   $x_3$: Petal Length (cm)
-   $x_4$: Petal Width (cm)

The data matrix $\mathcal{X}$ is of size $150 \times 4$. Our goal is to reduce this to a lower-dimensional representation while preserving as much variance as possible.

> **Note:** The scale of all variables should be the same for PCA to give meaningful results. Here all variables are in centimetres, so we do not scale.

```{r load-iris}
data(iris)
X <- as.matrix(iris[, 1:4])
head(X)
```

## Step-by-Step Manual PCA

### Step 1: Compute the Sample Mean

The sample mean vector $\bar{x}^\top$ is computed as:

$$\bar{x}^\top = \frac{1}{n}\sum_{i=1}^n x_i^\top$$

```{r mean}
mu <- colMeans(X)
mu
```

### Step 2: Center the Data

We subtract the mean from each observation to obtain the centred data matrix:

$$\mathcal{X}_c = \mathcal{X} - \mathbf{1}_n \bar{x}^\top$$

```{r center}
Xc <- sweep(X, 2, mu)
head(Xc)
```

### Step 3: Compute the Sample Covariance Matrix

$$S = \frac{1}{n} \mathcal{X}_c^\top \mathcal{X}_c$$

```{r covariance}
S <- t(Xc) %*% Xc / nrow(X)
round(S, 4)
```

### Step 4: Eigen Decomposition

The PCA solution requires decomposing $S$ into its eigenvalues and eigenvectors:

$$S = G \Lambda G^\top$$

where $\Lambda = \text{diag}(\lambda_1, \lambda_2, \ldots, \lambda_p)$ with $\lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_p$.

```{r eigen}
eig     <- eigen(S)
values  <- eig$values    # eigenvalues
vectors <- eig$vectors   # eigenvector matrix G

cat("Eigenvalues:\n")
round(values, 5)

cat("\nEigenvector matrix G:\n")
round(vectors, 4)
```

### Step 5: PCA Transformation

The scores (principal components) are obtained by projecting the centred data onto the eigenvectors:

$$\mathcal{Y} = \mathcal{X}_c \, G$$

```{r scores}
Y   <- Xc %*% vectors
Y_2 <- Y[, 1:2]          # keep first two PCs

cat("First 10 rows of the full transformed matrix Y (150×4):\n")
round(head(Y, 10), 4)

cat("\nFirst 10 rows of the reduced matrix Y(2) (150×2):\n")
round(head(Y_2, 10), 4)
```

The first two principal components capture the majority of the variance, so we retain the $150 \times 2$ matrix $\mathcal{Y}_{(2)}$ for visualisation and further analysis.

## Verification Using `prcomp`

```{r prcomp}
pca          <- prcomp(X, center = TRUE, scale. = FALSE)
Y_2_prcomp   <- pca$x[, 1:2]

cat("First 10 rows of Y(2) via prcomp:\n")
round(head(Y_2_prcomp, 10), 4)
```

The scores from the manual computation and `prcomp` agree up to possible sign flips (signs of eigenvectors are arbitrary).

## Visualisation: PCA Scatter Plot

Projecting onto the first two PCs reduces the data from 4 dimensions to 2, making class structure visible.

```{r pca-plot, fig.cap="PCA of the Iris Dataset: PC1 vs PC2 coloured by species."}
plot(Y_2[, 1], Y_2[, 2],
     col  = as.integer(iris$Species) + 1,
     pch  = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA of Iris Dataset")

legend("topright",
       legend = levels(iris$Species),
       col    = 2:4,
       pch    = 19,
       bty    = "n")
```

The plot shows that *setosa* is linearly separable from the other two species along PC1, while *versicolor* and *virginica* overlap somewhat along PC2.

------------------------------------------------------------------------

# Interpreting the Principal Components

## Eigenvector Weights

The first two PCs are linear combinations of the original variables:

$$y_1 = 0.361\,x_1 - 0.085\,x_2 + 0.857\,x_3 + 0.358\,x_4$$

$$y_2 = 0.657\,x_1 + 0.730\,x_2 - 0.173\,x_3 - 0.075\,x_4$$

```{r weights}
cat("Loadings (eigenvectors) for PC1 and PC2:\n")
round(vectors[, 1:2], 3)
```

## Proportion of Variance Explained and Scree Table

The proportion of total variance explained by the first $q$ PCs is:

$$\psi_q = \frac{\sum_{j=1}^q \lambda_j}{\sum_{j=1}^p \lambda_j}$$

```{r scree-table}
var_exp <- values / sum(values) * 100
cum_var <- cumsum(var_exp)

scree_table <- data.frame(
  PC          = 1:length(values),
  Variance_pct  = round(var_exp, 2),
  Cumulative_pct = round(cum_var, 2)
)

knitr::kable(scree_table,
             col.names = c("PC", "Var (%)", "Cum. (%)"),
             caption   = "Variance explained by each principal component (Iris).")
```

The first two PCs together explain **`r round(cum_var[2], 1)`%** of the total variation, confirming that the 2-dimensional projection is a highly faithful representation of the original 4-dimensional data.

## Scree Plot

```{r scree-plot, fig.cap="Scree plot: variance explained (%) by each PC for the Iris dataset."}
plot(var_exp, type = "b", pch = 19,
     xlab = "Principal Component",
     ylab = "Variance Explained (%)",
     main = "Scree Plot",
     col  = "steelblue")
abline(h = 0, lty = 2, col = "gray")
```

## Correlation Between Original Variables and PCs

To measure how strongly each original variable $X_i$ is related to each PC $Y_j$, we compute:

$$r_{X_i Y_j} = g_{ij} \left(\frac{l_j}{s_{X_i X_i}}\right)^{1/2}$$

where $g_{ij}$ is the $(i,j)$ entry of the eigenvector matrix, $l_j$ is the $j$-th eigenvalue, and $s_{X_i X_i}$ is the sample variance of $X_i$.

```{r cor-matrix}
var_X  <- diag(S)
var_Y  <- values
cov_XY <- S %*% vectors

cor_XY <- matrix(NA, nrow = ncol(X), ncol = ncol(X))
for (j in 1:ncol(X)) {
  for (k in 1:ncol(X)) {
    if (var_Y[k] > 1e-10)
      cor_XY[j, k] <- cov_XY[j, k] / sqrt(var_X[j] * var_Y[k])
  }
}
colnames(cor_XY) <- paste0("PC", 1:ncol(X))
rownames(cor_XY) <- c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")

knitr::kable(round(cor_XY, 3),
             caption = "Correlation between original variables and PCs (Iris).")
```

## Correlation Circle

The correlation circle plots $r_{X_i Y_1}$ against $r_{X_i Y_2}$ for each original variable. All points lie inside (or on) the unit circle because $r^2_{X_i Y_1} + r^2_{X_i Y_2} \leq 1$. Variables near the boundary are *well-explained* by the two-dimensional projection.

```{r cor-circle, fig.cap="Correlation circle: relationship between original Iris variables and the first two PCs."}
cor_XY_2 <- cor_XY[, 1:2]

plot(cor_XY_2[, 1], cor_XY_2[, 2],
     xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1),
     xlab = "PC1", ylab = "PC2",
     main = "Correlation Circle",
     pch  = 19, col = "darkred")

text(cor_XY_2[, 1], cor_XY_2[, 2],
     labels = rownames(cor_XY_2),
     pos    = 3, cex = 0.8)

theta <- seq(0, 2 * pi, length = 200)
lines(cos(theta), sin(theta), col = "blue", lwd = 1.5)
abline(h = 0, v = 0, col = "gray", lty = 2)
```

------------------------------------------------------------------------

# PCA on the Handwritten Digits Dataset

## Dataset Description

We use the **ZIP code handwritten digit** dataset (`ElemStatLearn` package). Each image is a $16 \times 16$ greyscale pixel grid, giving a **256-dimensional** feature vector per image. We focus on the **650 images of the digit 3**.

```{r load-digits}
# Install from local archive if needed, otherwise load the package
# install.packages("ElemStatLearn", repos = NULL, type = "source")
library(ElemStatLearn)

data(zip.train)
X_full <- as.matrix(zip.train[, -1])
y      <- zip.train[, 1]

threes <- X_full[y == 3, ]

cat("Dimensions of the digit-3 data matrix:", dim(threes), "\n")
cat("(", nrow(threes), "images ×", ncol(threes), "pixels )\n")
```

## Visualising a Sample of Original Images

```{r show-originals, fig.cap="Sample of 25 original handwritten digit-3 images.", fig.width=5, fig.height=5}
par(mfrow = c(5, 5), mar = c(0, 0, 0, 0))
for (i in 1:25) {
  image(matrix(threes[i, ], 16, 16)[, 16:1],
        col  = colorRampPalette(c("white", "red"))(256),
        axes = FALSE)
  box(col = "red")
}
par(mfrow = c(1, 1))
```

## Applying PCA

The PCA steps mirror those for Iris, but now $p = 256$.

```{r pca-digits}
pca_d   <- prcomp(threes, center = TRUE, scale. = FALSE)

X_bar_vec <- colMeans(threes)
n <- nrow(threes)
p <- ncol(threes)

X_bar <- matrix(X_bar_vec, nrow = n, ncol = p, byrow = TRUE)

G <- pca_d$rotation   # eigenvector matrix (256 × 256)
Y <- pca_d$x          # score matrix (n × 256)
```

## Image Reconstruction

Given the PCA transformation $\mathcal{Y} = \mathcal{X}_c \, G$, we can reconstruct the data using only $q$ principal components:

$$\hat{\mathcal{X}}_{(q)} = \mathbf{1}_n \bar{x}^\top + \mathcal{Y}_{(q)} \, G_{(q)}^\top$$

We compare reconstructions with $q = 2$ and $q = 10$ PCs against the original image.

```{r reconstruct}
# 2 PCs
Y_2  <- Y[, 1:2];   G_2  <- G[, 1:2]
X_hat_2  <- X_bar + Y_2  %*% t(G_2)

# 10 PCs
Y_10 <- Y[, 1:10];  G_10 <- G[, 1:10]
X_hat_10 <- X_bar + Y_10 %*% t(G_10)
```

```{r recon-plot, fig.cap="Reconstruction of the first digit-3 image using 2 and 10 PCs.", fig.width=7, fig.height=3}
par(mfrow = c(1, 3), mar = c(1, 1, 2, 1))

image(matrix(threes[1, ],   16, 16)[, 16:1],
      col  = colorRampPalette(c("white", "red"))(256),
      main = "Original", axes = FALSE)

image(matrix(X_hat_2[1, ],  16, 16)[, 16:1],
      col  = colorRampPalette(c("white", "red"))(256),
      main = "2 PCs", axes = FALSE)

image(matrix(X_hat_10[1, ], 16, 16)[, 16:1],
      col  = colorRampPalette(c("white", "red"))(256),
      main = "10 PCs", axes = FALSE)

par(mfrow = c(1, 1))
```

With just 10 PCs (out of 256), i.e. only **3.9%** of the original features, the reconstructed images are already recognisable as the digit 3.

```{r show-reconstructed, fig.cap="Sample of 25 reconstructed digit-3 images using 10 PCs.", fig.width=5, fig.height=5}
par(mfrow = c(5, 5), mar = c(0, 0, 0, 0))
for (i in 1:25) {
  image(matrix(X_hat_10[i, ], 16, 16)[, 16:1],
        col  = colorRampPalette(c("white", "red"))(256),
        axes = FALSE)
  box(col = "blue")
}
par(mfrow = c(1, 1))
```

------------------------------------------------------------------------

# Variance Explained vs. Number of Principal Components

## Cumulative Variance Table

```{r var-table}
lambda <- pca_d$sdev^2
psi    <- cumsum(lambda) / sum(lambda) * 100

qs <- c(1, 5, 10, 20, 50, 100, 256)
var_table <- data.frame(
  q                = qs,
  Cumulative_Var   = paste0(round(psi[qs], 1), "%"),
  Compression_Ratio = paste0(p, " : ", qs)
)

knitr::kable(var_table,
             col.names = c("q (# PCs)", "Cumulative Var. Explained", "Compression Ratio (p : q)"),
             caption   = "Cumulative variance explained at selected numbers of PCs (digit-3 dataset).")
```

## Reconstruction Error vs. Number of Principal Components

The total reconstruction error across all $n$ images when retaining $q$ components is, from the report's formula:

$$\sum_{i=1}^{n} \|\mathbf{x}_i - \hat{\mathbf{x}}_{i,(q)}\|^2 = (n-1)\sum_{j=q+1}^{p} \ell_j$$

We also report the **mean per-image error** (dividing by $n$) and the **mean per-pixel error** (dividing further by $p = 256$), which are easier to interpret on the original pixel-intensity scale.

```{r recon-error-table}
lambda <- pca_d$sdev^2
total_var <- sum(lambda)

qs <- c(1, 5, 10, 20, 50, 100, 150, 200, 256)

error_table <- data.frame(
  q                = qs,
  Total_SSE        = sapply(qs, function(q) (n - 1) * sum(lambda[seq(q + 1, p)])),
  Mean_Image_SSE   = sapply(qs, function(q) (n - 1) * sum(lambda[seq(q + 1, p)]) / n),
  Mean_Pixel_MSE   = sapply(qs, function(q) (n - 1) * sum(lambda[seq(q + 1, p)]) / (n * p)),
  Var_Retained_pct = round(cumsum(lambda)[qs] / total_var * 100, 1)
)

error_table[error_table$q == p,
            c("Total_SSE", "Mean_Image_SSE", "Mean_Pixel_MSE")] <- 0

error_table$Total_SSE      <- round(error_table$Total_SSE,      2)
error_table$Mean_Image_SSE <- round(error_table$Mean_Image_SSE, 4)
error_table$Mean_Pixel_MSE <- round(error_table$Mean_Pixel_MSE, 6)

knitr::kable(
  error_table,
  col.names = c(
    "q (PCs)",
    "Total SSE",
    "Mean SSE / image",
    "Mean MSE / pixel",
    "Var retained (%)"
  ),
  caption = "Reconstruction error at selected q for the digit-3 dataset."
)
```
